---
title: "HOWTO: Map Lat/Long to FIPS"
author: Andy Shi
date: '2020-11-02'
slug: mapping-lat-long-to-fips
categories:
  - R
tags:
  - HowTo
  - Spatial
---

# Introduction

A student in one of my classes had a bunch of latitude/longitude data from the
US and wanted to map them to FIPS codes, which are codes used to identify
individual counties in the US.

Let's first generate some fake data from Pennsylvania. We use the `us_counties`
function from `USAboundaries` to get
[`sf` shapefiles](https://cran.r-project.org/package=sf) of the counties.

```{r gen_data, message = FALSE}
library(sf)
library(dplyr)
library(tigris)
library(USAboundaries)

county_maps <- us_counties(resolution = "high", states = "PA")
n <- 1e5
set.seed(7477263)
county_bbox <- st_bbox(county_maps)
# randomly sample longitude and latitude from PA's bounding box
latlong_dat <- data.frame(id = seq_len(n),
                          x = rnorm(n),
                          Longitude = runif(n, min = county_bbox["xmin"],
                                            max = county_bbox["xmax"]),
                          Latitude = runif(n, min = county_bbox["ymin"],
                                            max = county_bbox["ymax"]))
```

# Original Solution

The original solution was to use `call_geolocator_latlong` to get the census
tract using an online API, and then convert the census tract into county.
However, the API takes a while to respond, so this solution isn't scalable
for thousands of data points.

```{r call_geolocator}
system.time({
  tigris::call_geolocator_latlon(lat = latlong_dat$Latitude[1],
                                 lon = latlong_dat$Longitude[1])
})
```

# Improved Solution

We will use the `st_intersects` function from the `sf` library and the
county-level `sf` shapefiles to find in which county a given lat/long pair is
contained. To do that, we convert the latitude and longitude data into an `sf`
object. Note that if you have NA values for the latitude or longitude, you'll
need to filter them out.

```{r convert}
latlong_sf <- latlong_dat %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = st_crs(county_maps))
```

Then, we use the `st_intersects` function to find the counties which intersect
with the points. This returns a list, where the ith element is the row number of
`county_maps` which contains the ith lat/long pair.

We might get a message like:

```
although coordinates are longitude/latitude, st_intersects assumes that they are planar
```
I believe this can be ignored unless you're working with data near the polar
regions.

```{r intersect, message = FALSE}
system.time({
  intersected <- st_intersects(latlong_sf, county_maps)
})
```

We can see that this runs super quickly---it took the same time to process
all 10K lat/long pairs with `st_intersects` as it did to process 1 pair with
`call_geolocator_latlon`.

We can then add the FIPS as a column to our latitude/longitude data. We can
use the intersection to subset the `geoid` column of county_maps, which
is the FIPS code. If the intersection was NA, then the FIPS code will be an
empty string (note FIPS codes are stored as strings because sometimes they have
a leading zero).

```{r latlong_final}
latlong_final <- latlong_sf %>%
  mutate(intersection = as.integer(intersected),
         fips = if_else(is.na(intersection), "",
                        county_maps$geoid[intersection]))
head(latlong_final)
```

To check our solution, we can draw a plot of all the lat/long pairs from a
specific PA county.

```{r plot}
county_maps %>%
  select(geometry) %>%
  plot()
latlong_final %>%
  filter(fips == "42019") %>%
  select(geometry) %>%
  slice_sample(n = 10) %>%
  plot(add = TRUE, reset = FALSE, pch = 16, col = "red")

```
