---
title: "HOWTO: Map Lat/Long to FIPS"
author: Andy Shi
date: '2020-11-02'
slug: mapping-lat-long-to-fips
categories:
  - R
tags:
  - HowTo
  - Spatial
---

# Introduction

A student in one of my classes had a bunch of latitude/longitude data from the
US and wanted to map them to a US Census tract and its corresponding code (FIPS
code).

We can first download the Census tracts for each state from the [US Census
website](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html)
under the "Census Tracts" dropdown menu. Then, after unzipping the file, we can
read the shape file using `st_read` from the sf package.

```{r read_shape, message = FALSE}
library(sf)
library(dplyr)
census_tracts <- st_read("cb_2018_42_tract_500k.shp", quiet = TRUE)
```

Then, we can generate some fake lat/long data for Pennsylvania. 

```{r gen_data, message = FALSE}
n <- 1e5
set.seed(7477263)
county_bbox <- st_bbox(census_tracts)
# randomly sample longitude and latitude from PA's bounding box
latlong_dat <- data.frame(id = seq_len(n),
                          x = rnorm(n),
                          Longitude = runif(n, min = county_bbox["xmin"],
                                            max = county_bbox["xmax"]),
                          Latitude = runif(n, min = county_bbox["ymin"],
                                            max = county_bbox["ymax"]))
```

# Original Solution

The original solution was to use `call_geolocator_latlong` to get the block FIPS
using an online API, and then convert that into Census tract by taking the first
11 digits. However, the API takes a while to respond, so this solution isn't
scalable for thousands of data points.

```{r call_geolocator, message = FALSE}
library(tigris)
system.time({
  census_block <- tigris::call_geolocator_latlon(lat = latlong_dat$Latitude[1],
                                                 lon = latlong_dat$Longitude[1])
})
(census_tract <- substr(census_block, start = 1, stop = 11))
```

# Improved Solution

We will use the `st_intersects` function from the `sf` library and the
county-level `sf` shapefiles to find in which county a given lat/long pair is
contained. To do that, we convert the latitude and longitude data into an `sf`
object. Note that if you have NA values for the latitude or longitude, you'll
need to filter them out.

```{r convert}
latlong_sf <- latlong_dat %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = st_crs(census_tracts))
```

Then, we use the `st_intersects` function to find the counties which intersect
with the points. This returns a list, where the ith element is the row number of
`census_tracts` which contains the ith lat/long pair.

We might get a message like:

```
although coordinates are longitude/latitude, st_intersects assumes that they are planar
```
I believe this can be ignored unless you're working with data near the polar
regions.

```{r intersect, message = FALSE}
system.time({
  intersected <- st_intersects(latlong_sf, census_tracts)
})
```

We can see that this runs super quickly---it took the same time to process
all 10K lat/long pairs with `st_intersects` as it did to process 1 pair with
`call_geolocator_latlon`.

We can then add the FIPS as a column to our latitude/longitude data. We can use
the intersection to subset the `GEOID` column of `census_tracts`, which is the
FIPS code. If the intersection was NA, then the FIPS code will be an empty
string (note FIPS codes are stored as strings because sometimes they have a
leading zero).

```{r latlong_final}
latlong_final <- latlong_sf %>%
  mutate(intersection = as.integer(intersected),
         fips = if_else(is.na(intersection), "",
                        census_tracts$GEOID[intersection]))
head(latlong_final)
```

To check our solution, we can plot all the Census tracts in Philadelphia (county
code 42101) and display the points which showed up in a particular Census tract.

```{r plot}
census_tracts %>%
  filter(startsWith(GEOID, "42101")) %>%
  select(geometry) %>%
  plot()
latlong_final %>%
  filter(fips == "42101980300") %>%
  select(geometry) %>%
  slice_sample(n = 10) %>%
  plot(add = TRUE, reset = FALSE, pch = 16, col = "red")

```
